---
title: "Ascertain diagnoses"
description: >
  Learn how to ascertain diagnoses from all UK Biobank sources.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Ascertain diagnoses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ukbrapR)
```

Diagnosis of condition in UK Biobank participants come from multiple data sources:

* Self-report during assessment

* Hospital inpatient records (HES)

* Primary care (GP)

* Cancer registry

* Cause of death

The {ukbrapR} package makes it fast and easy to ascertain diagnoses from multiple UK Biobank data sources in the DNAnexus Research Analysis Platform (RAP).

Depending on the data source different coding vocabularies are required:

* `ICD10` (for searching HES diagnoses, cause of death, and cancer registry)

* `ICD9` (for searching older HES diagnosis data)

* `Read2` and `CTV3` (for GP clinical events)

* `OPCS3` and `OPCS4` (for HES operations)

* `ukb_cancer` and `ukb_noncancer` (for self-reported illness at UK Biobank assessments - all instances will be searched)

Ascertaining diagnoses typically takes two steps:

## 1. Get medical records and self-reported illness data for provided codes

For a given set of diagnostic codes get the participant medical events and self-reported data. Returns a list of 6 data frames: the subset of the long clinical files with matched codes. 

Codes need to be provided as a data frame with two fields: `vocab_id` and `code`. Valid code vocabularies are listed above. Other cols (such as condition and description) are ignored.

```r
# example diagnostic codes for Chronic Kidney Disease 
codes_df_ckd <- ukbrapR:::codes_df_ckd
head(codes_df_ckd)
#>   condition vocab_id  code
#> 1       ckd    ICD10 N18.3
#> 2       ckd    ICD10 N18.4
#> 3       ckd    ICD10 N18.5
#> 4       ckd    ICD10 N18.6
#> 5       ckd    ICD10 N18.9
#> 6       ckd    ICD10   N19

# get diagnosis data - returns list of data frames (one per source)
diagnosis_list <- get_diagnoses(codes_df_ckd) 

# N records for each source
nrow(diagnosis_list$gp_clinical)
nrow(diagnosis_list$hesin_diag)
nrow(diagnosis_list$death_cause)
```

## 2. Get date first diagnosed

Usually the user is interested in combining the separate data sources into a combined phenotype: the date first diagnosed for each participant from the data/codes in step 1 (cause of death, HES diagnoses, GP clinical, cancer registry, HES operations, and self-reported illness fields). 

In addition to the "date first" `df` field are:

 - a `src` field indicating the source of the date of first diagnosis.
 - a `bin` field indicating the cases [1] and controls [0]. This relies on a small number of baseline fields also exported. The `df` field for the controls is the date of censoring (currently 30 October 2022).
 - a `bin_prev` field indicating whether the case was before the UK Biobank baseline assessment

```r
# for each participant, get Date First diagnosed with the condition
diagnosis_df <- get_df(diagnosis_list)

names(diagnosis_df)
summary(diagnosis_df)
```

You can add a prefix to all the variable names by specifying the "prefix" option:

```r
diagnosis_df <- get_df(diagnosis_list, prefix="ckd")

# how many cases ascertained?
table(diagnosis_df$ckd_bin)

# source of earliest diagnosis date
table(diagnosis_df$ckd_src)

# date of diagnosis for prevalent cases (i.e., before UK Biobank baseline assessment)
summary(diagnosis_df$ckd_df[ diagnosis_df$ckd_bin_prev == 1 ])
```

## 3. Ascertaining multiple conditions at once 

The default `get_df()` behaviour is to use all available codes. However, the most time-efficient way to get multiple conditions is to run `get_diagnoses()` once for all codes for the conditions you wish to ascertain, then get the "date first diagnosed" for each condition separately. In the codes data frame you just need a field indicating the condition name, that will become the variable prefixes.

```r
# combine haemochromatosis and CKD codes together
#   each contain there columns: condition, vocab_id, and code
#   where `condition` is either "hh" or "ckd" and will become the variable prefix
codes_df_combined = rbind(ukbrapR:::codes_df_hh, ukbrapR:::codes_df_ckd)

# get diagnosis data - returns list of data frames (one per source)
diagnosis_list <- get_diagnoses(codes_df_combined)

# for each participant, get Date First diagnosed with the condition
diagnosis_df = get_df(diagnosis_list, group_by="condition")

# each condition has full set of output
table(diagnosis_df$hh_bin)

table(diagnosis_df$ckd_bin)
```

In the above example we also included a UK Biobank self-reported illness code for haemochromatosis, that was also ascertained (the Date First is run on each condition separately, they do not all need to have the same data sources).


